Forage site distributions quantify collective foraging in honey bee colonies

### 2022-01-27 analysis/manuscript/paper.Rmd
added refs to MS and updated a methods section

Honeybee foraging is a collective behaviour that is directed through the extraordinary waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops. Despite having been described as the most astounding example of non-primate communication, recent evidence suggests that the benefits of using dance information vary across environments and are sometimes even negligible. Yet, because we cannot easily quantify collective foraging, pinpointing which environments promote collective foraging is challenging. Here, we present a mathematical model which allows inferring the proportion of 'scout' and 'recruit' dances on a hive's dance floor and quantify collective foraging. By applying this methodology to waggle dance data from twenty hives, we show that dance information may be used less in environments where resources are sparse, potentially because colonies are required to send out more scouts to find food in these circumstances. Our methodology provides a tool to quantify waggle dance use in undisturbed hives at a much larger scale than hitherto has been possible and to identify those ecological conditions that may have driven the evolution of this remarkable collective behaviour.

Needs referencing.

@@

Within honeybee colonies, a series of simple rules that determine when and how much bees perform the celebrated waggle dance (Box 1) mean that choices between feeding sites occur at the level of the group rather than the individual. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the sweeter of two equidistant sources (ref), or the closer of two equally sweet sources (ref), without requiring any individual bee to compare options. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006].

@@

The prevalent explanation for this apparent redundancy portrays dance-based collective foraging as an evolutionary anachronism that was useful in ancient tropical environments but offers little benefit in modern temperate environments (ref). However, fossil evidence has questioned the long-held assumption that the genus Apis evolved in tropical climates (Kotthoff et al 2013). Moreover, given that dancing comes at a significant time cost (ref), this raises the issue of why all temperate *Apis mellifera* colonies continue to dance. The dual questions of what foraging selection pressures drove the evolution of this uniquely complex collective behaviour, and why it continues to exist, are thus yet to be satisfactorily answered. Empirical studies have shown that honeybee colonies use dance information flexibly in response to their environment, suggesting that bees cease to rely on dance information if it provides little benefit [@Price2019]. Thus, identifying those environments whereby modern honeybee colonies forage collectively rather than through individual search is a key step towards understanding dance evolution (ref).

@@

Honeybees provide an ideal study system in which to evaluate collective foraging thanks to the famous “waggle dance”, which conveys information about location of a resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) (von Frisch 1993). Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Resource “quality” is conveyed through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]). Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A].
---

Overall, evidence that colony-level collective foraging offers any benefit over individual search has proved surprisingly hard to come by, mainly because this requires to distinguish scouts from recruits. Scouts are foragers that have located a resource through individual search behaviour, whilst recruits are foragers that have found a forage site by using the information provided in the dance of another bee. Scouting and recruiting are not fixed behavioural categories, because individual bees can engage in both over the course of their foraging lifetime, and foragers can dance on return from any successful trip irrespective of whether they were recruited to the forage site or found it individually. So far, scouts and recruits are distinguished by means of the foragers' individual search history, which requires following individual bees and is time and labour intensive.

@@

To reveal what the distribution of waggle run durations reported on the dance floor look like, and to demonstrate that dances from recruits and scouts lead to different distributions, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor (Fig. 1). As it is known in the simulation which individuals forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions for the two types of foraging trips are very different. The distance distribution of the scouts is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) is a Rayleigh distribution which is the nearest-neighbour distribution of a two-dimensional environment [@Pyke1978] (Pyke 1978) (see Methods).

@@

Scouts and recruits create different foraging distance distributions on the dance-floor. These different distance distributions can be used to quantify the relative importance of collective foraging. Foragers that scout report a distribution of distances that reflects their probability of encountering a resource in the environment (see methods), whilst foragers that recruits report distances that reflect the distribution of the most profitable resource available (see methods). On the dance-floor, the distribution of distances is modified through the number of runs performed for a resource, which depends on the profitability of a resource, which honeybees achieve by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994A]. By combining this profitability with the distributions identified in our simulations, we develop a method to describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 2, see Methods).

This method intrinsically captures honeybee foraging as collective decision making as the foraging sites represented on the dance floor derive from a mixture of individual search and waggle dance information, captured in the proportions p and 1-p, respectively. Consequently, by setting p to 1, we can also derive a model based on the sole use of individual search (scouts only) (Fig 2). By fitting these two models to waggle run duration distributions from real honeybee colonies we are thus able, using model selection [@BurnhamAnderson?](Burnham and Anderson), to infer if and, by estimating the parameter p, to what extent honeybee colonies use social information when foraging.

@@

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed an existing dataset of 3378 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives were situated at different locations in South East England (see methods, figure 3A) and were visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually ([@Couvillon2012; @Schurch2019] to extract waggle run durations. For each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter p. We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data.

@@

Our results show that, whilst colony-level foraging mostly use a mixture of scouts and recruits, in some circumstances, colony foraging is best explained by a model of individual foraging. This suggests that in some environments, honey bees actively ignore the waggle dance and instead prioritise the use of private information. Previous empirical studies have shown that individual foragers flexibly alter their use of waggle dance information in response to unrewarding resources [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individuals foraging behaviour over a landscape. As our results show that these individual behaviours accurately describe colony foraging in three different sites, these findings support the idea that these individual behaviours can dominate colony foraging.

Our finding that foraging in some landscapes is more accurately described by ignoring the dance and focusing on private information supports the hypothesis that the dance is used flexibly in response to environmental factors. As the collective model quantifies waggle dance use, the fit of our model provides a tool to evaluate how environmental factors influence waggle dance use, and thus how a colony values social and private information in different environmental circumstances. This allows for comparing waggle dance use across environments.

We compared the estimates of the proportion of scout and recruits for colonies in different environments. Each site used in this study is situated in a landscape comprised of different proportions of land-use types. To evaluate how these different land-use types influenced waggle dance use, we first classified the different land-use types of the area surrounding each site to obtain a standardised land-use profile for the urban and agri-rural environments separately (see methods). As many of the resulting land-use types covaried with each other, we performed a Partial Least Squares (PLS) analysis to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances.

@@

In agri-rural environments, land-use types which typically associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with a higher resource abundance correlate with a lower proportion of scouts. Arable land is typically nutritionally poor for honey bees (ref). On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas in agri-rural environments, which often have gardens with ample flowers available for pollinators (ref). These findings indicate that in agri-rural environments, when resources are more abundant and thus easier to find, the collective foraging is used more, presumably to identify the best resources in the environment (ref). Conversely, where resources are more difficult to find, a decreased use of the waggle dance suggests the colony focuses more on individual exploration rather than in trying to locate on the best resources. A similar result is observed for urban environments, where the dependence appears to be driven by the presence of flower rich gardens in residential areas. However, an increase in the amount of parks, allotments and cemeteries decreases the use of the dance.

>>>>

Honeybee foraging is a collective behaviour that is directed through the extraordinary waggle dance communication system, whereby colony foraging effort is allocated through a series of feedback loops [@Frisch1993; @Seeley1995]. Despite having been described as the most astounding example of non-primate communication [@Frisch1993], recent evidence suggests that the benefits of using dance information vary across environments and are sometimes even negligible [@Dornhaus2006; @Price2019]. Yet, because we cannot easily quantify collective foraging, pinpointing which environments promote collective foraging is challenging. Here, we present a mathematical model which allows inferring the proportion of 'scout' and 'recruit' dances on a hive's dance floor and quantify collective foraging. By applying this methodology to waggle dance data from twenty hives, we show that dance information may be used less in environments where resources are sparse, potentially because colonies are required to send out more scouts to find food in these circumstances. Our methodology provides a tool to quantify waggle dance use in undisturbed hives at a much larger scale than hitherto has been possible and to identify those ecological conditions that may have driven the evolution of this remarkable collective behaviour.

@@

Within honeybee colonies, a series of simple rules that determine when and how much bees perform the celebrated waggle dance (Box 1) mean that choices between feeding sites occur at the level of the group rather than the individual. For example, because the number of dance circuits performed by a bee on returning from a food source reflects the net energetic benefits of the trip, more of the colony's workforce will be recruited to the sweeter of two equidistant sources [@Seeley1991], or the closer of two equally sweet sources [@Hasenjager2022], without requiring any individual bee to compare options. This extraordinary system is a key example of how social insect colony behaviour can take on a form extending beyond that of the individual units [@Sumpter2006].

@@

The prevalent explanation for this apparent redundancy portrays dance-based collective foraging as an evolutionary anachronism that was useful in ancient tropical environments but offers little benefit in modern temperate environments [@Dornhaus2004; @Donaldson-Matasci2012; @Price2015]. However, fossil evidence has questioned the long-held assumption that the genus Apis evolved in tropical climates [@Kotthoff2013]. Moreover, given that dancing comes at a significant time cost [@Seeley1988], this raises the issue of why all temperate *Apis mellifera* colonies continue to dance. The dual questions of what foraging selection pressures drove the evolution of this uniquely complex collective behaviour, and why it continues to exist, are thus yet to be satisfactorily answered. Empirical studies have shown that honeybee colonies use dance information flexibly in response to their environment [@Sherman2002; @Dornhaus2004; @Dornhaus2006], suggesting that bees cease to rely on dance information if it provides little benefit [@Price2019]. Thus, identifying those environments whereby modern honeybee colonies forage collectively rather than through individual search is a key step towards understanding dance evolution [@Price2015].

@@

Honeybees provide an ideal study system in which to evaluate collective foraging thanks to the famous “waggle dance”, which conveys information about location of a resource as a vector of distance (duration of the waggle run) and direction (angle of the dance) [@Frisch1993]. Through the observation and decoding of the waggle dance, a colony's dance floor provides a unique opportunity to eavesdrop on the communication and decision making leading to collective foraging decisions. Resource “quality” is conveyed through the number of waggle runs performed [@Frisch1993; @Boch1956; @Esch1961; @Seeley1994; @Seeley2000]). Although bees that follow dances do not specifically interpret this information on an individual level [@Hasenjager2021], the resulting over-representation of high-quality sites on the dance-floor means that they are more likely to encounter dances that advertise better forage [@Seeley1995], and provides the colony with a mechanism to select the most profitable resources in their environment [@Seeley1994A].

---

Overall, evidence that colony-level collective foraging offers any benefit over individual search has proved surprisingly hard to come by, mainly because this requires to distinguish scouts from recruits. Scouts are foragers that have located a resource through individual search behaviour, whilst recruits are foragers that have found a forage site by using the information provided in the dance of another bee [@Seeley1995]. Scouting and recruiting are not fixed behavioural categories, because individual bees can engage in both over the course of their foraging lifetime, and foragers can dance on return from any successful trip irrespective of whether they were recruited to the forage site or found it individually [@Beekman2007]. So far, scouts and recruits are distinguished by means of the foragers' individual search history, which requires following individual bees and is time and labour intensive [@Seeley1995].

@@

To reveal what the distribution of waggle run durations reported on the dance floor look like, and to demonstrate that dances from recruits and scouts lead to different distributions, we simulated honeybees foraging in a landscape where resource patches were randomly placed in the environment. Foragers could locate these under two different strategies: either acting as a scout and locating resources themselves, or following a recruit strategy and locating resources by following a random dance from the dance floor [@Seeley1992] (Fig. 1). As it is known in the simulation which individuals forage under what strategy, we can compare the distributions of foraging distances reported on the dance floor. Figs 1c and 1d show that the shapes of the resource distance distributions for the two types of foraging trips are very different. The distance distribution of the scouts is akin to that of an exponential distribution (Fig. 1 C), which is the nearest neighbour distance distribution for foragers operating in a one-dimensional environment (see methods). The distribution of the distances reported by recruits (Fig. 1 D) is a Rayleigh distribution which is the nearest-neighbour distribution of a two-dimensional environment [@Pyke1978] (see Methods).

@@

Scouts and recruits create different foraging distance distributions on the dance-floor. These different distance distributions can be used to quantify the relative importance of collective foraging. Scouting foragers report a distribution of distances that reflects their probability of encountering a resource in the environment (see methods), whilst foragers that recruits report distances that reflect the distribution of the most profitable resource available (see methods). On the dance-floor, the distribution of distances is modified through the number of runs performed for a resource, which depends on the profitability of a resource, which honeybees achieve by measuring the energetic efficiency of a foraging trip through the ratio of energetic gain to energetic cost [@Seeley1994A]. By combining this profitability with the distributions identified in our simulations, we develop a novel method to describe the distribution of waggle runs reported on the dance-floors of real honeybee colonies as a superposition of scout and recruit distributions (Fig 2, see Methods).

This method intrinsically captures honeybee foraging as collective decision making, as the foraging sites represented on the dance floor derive from a mixture of individual search and waggle dance information, captured in the proportions $p$ and $1-p$, respectively. Consequently, by setting $p = 1$, we can also derive a model based on the sole use of individual search (scouts only) (Fig 2). By fitting these two models to distribution of waggle run durations decoded from real honeybee colonies we are thus able, using model selection [@Burnham2002], to infer if and, by estimating the parameter $p$, to what extent honeybee colonies use waggle dance information when foraging.

@@

To evaluate the use of waggle dance information and individual search in honeybee colonies foraging in 'natural' landscapes, we analysed an existing dataset of 3378 waggle dance observations from 20 observation hives, recorded between April-September 2017, (previously described in [@Samuelson2021]). Hives were situated at different locations in South East England (see methods, figure 3A) and were visited every two weeks for a period of 24 weeks. On each visit, two hours of continuous waggle dance data was recorded by training a camcorder onto the dance floor. The footage of the dances was decoded manually ([@Couvillon2012; @Schurch2019] to extract waggle run durations. For each site we fit both the collective and individual models and used model selection to determine which provided the better explanation of the data, and (if the collective model provided a better fit) to quantify the relative use of social information through estimating the parameter p. We calculated the goodness-of-fit using a Kolmorgorov-Smirnov (KS) test to ascertain if the model provided a plausible explanation of the data [@Goldstein2004].

@@

Our results show that, whilst colony-level foraging mostly use a mixture of scouts and recruits, in some circumstances, colony foraging is better described by a model of individual foraging. This suggests that in some environments, honey bees actively ignore the waggle dance and instead prioritise the use of private information. Previous empirical studies have shown that individual foragers flexibly alter their use of waggle dance information in response to unrewarding resources [@Gruter2011; @Price2019]. Evaluating how these individual decisions influence the collective, however, has historically been a challenge as it is effectively impossible to track an individuals foraging behaviour over a landscape. As our results show that individual foraging accurately describes colony foraging in three different sites, these findings support the idea that individual decision making can dominate colony foraging and demonstrates further evidence for flexible waggle dance use by honey bee colonies.

Our finding that foraging in some landscapes is more accurately described by ignoring the dance and focusing on private information supports the hypothesis that the dance is used flexibly in response to environmental factors. As the collective model quantifies waggle dance use, the fit of our model provides a new tool to evaluate how environmental factors influence waggle dance use, and thus how a colony values waggle dance information under different environmental circumstances.

We compared the estimates of the proportion of scout and recruits for colonies in different environments. Each site used in this study is situated in a landscape comprised of different proportions of land-use types. To evaluate how these different land-use types influenced waggle dance use, we first classified the different land-use types of the area surrounding each site to obtain a standardised land-use profile for the urban and agri-rural environments separately (see methods). As many of the resulting land-use types covaried with each other, we performed a Partial Least Squares (PLS) analysis [@Carrascal2009] (see methods) to determine the principal components that represent combinations of land-use types which explained the most variation in the proportion of scout dances.

@@

In agri-rural environments, land-use types which typically associate with resource scarcity correlate positively with a higher proportion of scouts, whilst those which correlate with a higher resource abundance correlate with a lower proportion of scouts. Arable land is typically nutritionally poor for honey bees [@Samuelson2021]. On the contrary, non-agricultural unmanaged green space is typically nutrient rich, as are built up areas, which often have gardens with ample flowers available for pollinators [@Baldock2019]. These findings indicate that in agri-rural environments, when resources are more abundant and thus easier to find, the collective foraging is used more, presumably to identify the best resources in the environment. Conversely, where resources are sparse and more difficult to find, a decreased use of the waggle dance suggests the colony focuses more on individual exploration rather than in trying to locate on the best resources. A similar result is observed for urban environments, where the dependence appears to be driven by the presence of flower rich gardens in residential areas. However, an increase in the amount of parks, allotments and cemeteries appears to decrease the use of the dance.

( **NOTE:** I feel like we need more of an explanation here. For the agri-rural environment we can assemble the results into a coherent story which has some logical basis and suggests more data would investigate this further. The results for the urban sites are contradictory. Parks, allotments and cemeteries increases the proportion of scouts so following our logic these areas should be barron, but the evidence suggests they are not and along side gardens should be full of flora. This indicates there is something else in play here, such as the spatial arrangement of the locations? Or that we just don't have enough data to comment on what processes are going on in urban sites. However, we do get a 'significant' result in the PLS. Maybe we need more fine grained data to evaluate how season influences, it could be that these results are muddled by an interaction with season or temperature or something. Also, railway is a random thing to pop up as significant. I am not sure how best to describe these results without it looking like blatant data mining...)
